### 2025-04-21 sec/0_abstract.tex
Some polishing

\begin{abstract}
Outlier-robust point cloud registration where correspondences are given but up to 90\% of them may be outliers is commonly formulated as as the \textit{Maximum Consensus} (MC) or \textit{Truncated Least Squares} (TLS) optimization problem.
This formulation has been shown to be very robust to outliers but actually solving the underlying combinatorial optimization problem exactly, i.e. with global optimality guarantee, remains challenging. Approaches using semidefinite relaxations are not well scalable albeit running in polynomial time and instead have a runtime on the order of hundreds of seconds for 100 points. O

We propose a new convex relaxation that runs in linear time and can be used in combination with Branch-and-Bound. We present results on synchetic data for the rotation-only (Wahba) problem and translation problem separately.

>>>>

\begin{abstract}
Outlier-robust point cloud registration where correspondences are given but up to 90\% of them may be outliers is commonly formulated as as the \textit{Maximum Consensus} (MC) or \textit{Truncated Least Squares} (TLS) optimization problem.
This formulation has been shown to be very robust to outliers but actually solving the underlying combinatorial optimization problem exactly, i.e. with global optimality guarantee, remains challenging. Approaches using semidefinite convex relaxations albeit running in polynomial time do not scale well and instead have a runtime on the order of hundreds of seconds for 100 points.

We propose a new convex relaxation that runs in linear time and can be used in combination with Branch-and-Bound. It is two orders of magnitude faster when solving the rotation-only problem compared to state-of-the-art SDP solver STRIDE. 
We present results on synthetic data for the rotation-only (Wahba) problem and translation problem separately.

### 2025-04-21 sec/3_tls_opt.tex
Some polishing

where $\mathbf{x}_i$ are data points and $r(\cdot)^2$ is some convex residual function such as the squared Euclidean distance $||\mathbf{x} - \mathbf{x}_i||_2^2$. In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas). Also, sums of convex functions are convex (this is the reason why ordinary least squares is a convex problem), so we can focus on making each term convex separately.

@@

However, some additional special case handling is required to ensure that this relaxation becomes tight when computing over (small) intervals where the TLS objective is already convex. This is crucial to ensure that BnB converges in a finite number of iterations. First, the reweighting should only be applied to residuals where $r_{min} < \epsilon^2$, i.e., residuals that are not already outliers over the entire interval. Similarly, residuals where $r_{max} < \epsilon^2$ are inliers over the entire interval and are therefore already convex and do not need to be reweighted.

After considering these special cases, we can show that $\text{WLS}_{r}$ is a convex relaxation of \ref{eq:tls-scalar-for-wls} over a given interval:
\begin{equation}
	\label{eq:tls-wls-relaxation1}
	\begin{aligned}	
		\text{WLS}_{r} &= 
		\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
	\end{aligned}
\end{equation}

@@

The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists \cite{Kabsch-1978-Point-set-alignment} \cite{Lawrence2019APA} \cite{Least-squares-estimation-point-sets-Umeyama-1991}. This problem however is modified and has additionally an inequality-constraint. We show that this problem can be solved computationally very efficient using a custom active-set solver.

The idea of the active set method is based on the observation that every inequality constraint can either be ignored or is satisfied as an equality (also called complementary slackness condition) \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}. We therefore first solve the problem using the SVD algorithm and thereby ignoring the constraint. Then, we simply check whether the constraint is satisfied. If yes, we already have found the solution. If not, we know that we need to find a 3D rotation that has a rotation angle equal to $n_r$. By the Euler theorem, every 3D rotation can be described as a rotation around a single axis and an single angle. Because we know that the rotation angle needs to be $n_r$, we therefore only need to find the rotation axis.

@@

Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient. The only computationally intensive operation is computing the $3 \times 3$ cross-correlation matrix that is needed first by the SVD algorithm. It has a linear runtime in the number of points. All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e. independent of the number of points). (For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points.)
This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

>>>>

where $\mathbf{x}_i$ are data points and $r(\cdot)^2$ is some convex residual function such as the squared Euclidean distance $||\mathbf{x} - \mathbf{x}_i||_2^2$. In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas). Since sums of convex functions are convex, we can focus on relaxing each term separately.

@@

Some additional case handling is required however to ensure that this relaxation becomes tight. This is crucial to ensure that BnB converges in a finite number of iterations. First, all residuals for which  $r_{min} \geq \epsilon^2$ holds are outliers over the entire interval and therefore constant. Similarly, residuals where $r_{max} \leq \epsilon^2$ holds are inliers over the entire interval and do not need to be reweighted.

After considering these special cases, we can show that $\text{WLS}_{relax}$ is a convex relaxation of \ref{eq:tls-scalar-for-wls} over a given interval:
\begin{equation}
	\label{eq:tls-wls-relaxation1}
	\begin{aligned}	
		\text{WLS}_{relax} &= 
		\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
	\end{aligned}
\end{equation}

@@

The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists \cite{Kabsch-1978-Point-set-alignment} \cite{Lawrence2019APA} \cite{Least-squares-estimation-point-sets-Umeyama-1991}. Our problem however has additionally an inequality-constraint. It still can be solved computationally very efficient using a custom active-set solver.

The idea of the active set method is based on the observation that every inequality constraint can either be ignored or is satisfied as an equality (also called complementary slackness condition) \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}. We therefore first solve the problem using the SVD algorithm and thereby ignore the constraint. Then, we simply check whether the constraint is satisfied. If yes, we already have found the solution. If not, we know that we need to find a 3D rotation that has a rotation angle equal to $n_r$. By the Euler theorem, every 3D rotation can be described as a rotation around a single axis and an single angle. Because we know that the rotation angle needs to be $n_r$, we therefore only need to find the rotation axis.

Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient. The only computationally intensive operation is computing the $3 \times 3$ cross-correlation matrix that is needed first by the SVD algorithm. It has a linear runtime in the number of points. All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e. independent of the number of points).
This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

@@

### 2025-04-21 sec/4_evaluation.tex
Some polishing

\section{Experimental evaluation}
\label{sec:evaluation}
In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
We measure the required number of iterations as well as the per-iteration costs. We have implemented a simple version of a best-first Branch-and-Bound \cite{10.1007/978-3-642-37444-9_42}.

@@

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
	\end{adjustbox}
	\caption{Der durchschnittliche Rotationsfehler in Abhängigkeit vom Anteil der Ausreißer: Unsere Methode bleibt robust bis zu 95\% Ausreißer, wie von der Truncated Least Squares Formulierung zu erwarten ist.}
	\label{fig:rotation-error-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
	\end{adjustbox}
	\caption{Die durchschnittliche Ausführungszeit von Branch-and-Bound in Abhängigkeit von der Gesamtzahl der Punkte $N$ des Optimierungsproblems.}
	\label{fig:runtime}
\end{figure}

>>>>

\section{Experimental evaluation}
\label{sec:evaluation}
In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
We measure the required number of iterations as well as the per-iteration costs. For this we have implemented a simple version of a best-first Branch-and-Bound as in \cite{10.1007/978-3-642-37444-9_42}.

@@

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
	\end{adjustbox}
	\caption{Rotation-only experiments: The average rotation error over the ourlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
	\label{fig:rotation-error-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
	\end{adjustbox}
	\caption{Average runtime of our Branch-and-Bound solver depending on the total amount of point correspondences $N$.}
	\label{fig:runtime}
\end{figure}
